How Believing in Free Will is Enough to Be More Satisfied at Work

### 2024-02-17 outputs/paper/paper.qmd

Utilizing the same dataset and a similar methodology to @feldman2017, we aim to replicate their results in the context of the United States, a very heterogenous country that is characterized by a high importance of individualist, but also cultural diversity and socioeconomic inequality. While Feldman et al. did investigate the relationship between the belief in free will and job satisfaction in the United States, they did not integrate these results to the cross-cultural analysis, as the US were not included in the sample of 16 countries. This paper, therefore, seeks to fill this gap and reproduce these aspects of the Feldman et al. paper: (1) the relationship between the belief in free will and job satisfaction applies to the context of American adults working at Amazon Mechanical Turk (2) free will is a significant predictor of job satisfaction at the national-level, using the World Values Survey. We will also consider whether there are any significant differences between the national-level data and the Amazon Mechanical Turk data. The estimand is the correlation between the belief in free will and job satisfaction. Utilizing Amazon Mechanical Turk and World Values Survey data, we found that belief in free will significantly correlates with job satisfaction both at individual and national levels. Additionally, we found job autonomy as a critical mediator in this relationship, while other agency constructs and demographic factors play a less important role in impacting workplace satisfaction across diverse American contexts. We also found American participants to be among the world's highest believers in free will, further emphasizing the importance of considering the local context.

We begin our paper with a closer look at the data source and measurement (@sec-data), followed by a selection of relevant results (@sec-results). Finally, we will conclude with a discussion of our findings to integrate our replication with the original paper by @feldman2017 and the broader literature on free will beliefs (@sec-discussion). We will also discuss potential ethical biases, limitations, and future directions. Our reproduction was conducted using the statistical programming language R [@citeR]. In the data analysis and visualization process, we also made use of the following R packages: tidyverse [@citetidyverse], dplyr [@citedplyr], knitr [@citeknitr], kableExtra [@citekExtra], purrr [@citepurrr], psych [@citepsych], car [@citecar], MASS [@citeMASS], and countrycode [@citecc].

@@

* Pride in Job: We expect this variable to positively correlate with job satisfaction and job autonomy, such that it can serve as some indication of internal validity and reliability. Not all waves of WVS data asked this question, but where possible we used the following question (C031):
  + How much pride, if any, do you take in the work that you do? (1 = "A great deal", 3 = "Little").
  + We also inverted this scale, such that 1 means "Little" and 3 means "A great deal", making it making it more consistent with the ordering logic of other variables used.

@@

kable(reliability_scores)
```
\newpage
## Data Characteristics {#sec-characteristics}

@@

kable(satisfaction_data_usa_summary)
```

@@

kable(wvs_data_usa_summary)

@@

In exploring free will beliefs and their impact on job satisfaction in the context of American Amazon Mechanical Turk workers, our study replicates and extends Feldman et al.'s results. As seen in @fig-corst2, most variables are significantly correlated with each other, with the highest correlations being unsurprisingly between the two measurements of Job satisfaction. Locus of control lacks any significant correlations, which together with its low reliability coefficient designates it as a less decisive control variable. The most important relationship, however, is between free will beliefs and job satisfaction: The belief in free will was significantly positively correlated with job satisfaction both at Time 1 (*r* = `r cor_fw_js1_correlation_coefficient`, *p* \<.001, CI \[`r cor_fw_js1_ci[1]`, `r cor_fw_js1_ci[2]`\]) and at Time 2 (r = `r cor_fw_js2_correlation_coefficient`, *p* \<.001, CI \[`r cor_fw_js2_ci[1]`, `r cor_fw_js1_ci[2]`\]). However, this correlation is not as strong as the one between free will beliefs and job autonomy (*r* = `r cor_fw_jaut1_correlation_coefficient`, *p* \<.001, CI \[`r cor_fw_jaut1_ci[1]`, `r cor_fw_jaut1_ci[2]`\]). This supports the claim that job autonomy may be the most important mediating variable between free will beliefs and job satisfaction. Moreover, a step-wise multiple regression on job satisfaction with all psychological agency constructs (full results in @sec-apx-stepwise) confirmed that the effect of free will beliefs on job satisfaction is robust and the strongest relative impact, even when controlling for other agency constructs (Time 1: *p* \< 0.034; Time 2: *p* \< 0.031).

@@

ggplot(satisfaction_data_usa_rel, aes(x = factor(Sex, labels = c("Male", "Female")), y = `Belief in Free Will (T1)`)) +
  geom_boxplot(fill = "lightblue") +
  labs(
       x = "Sex",
       y = "Belief in Free Will") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  + theme_minimal()

@@

## Effect of Free Will Beliefs
The main result of our analysis is that the belief in free will is sufficient to positively influence one's job satisfaction. Both the Amazon Mechanical Turk and WVS data sources showed significant positive correlations between these two psychological constructs. Free will beliefs are found to be  integral to psychological well-being and motivation, and are commonly linked to agency, autonomy, and responsibility [@wang2009]. By fostering an environment that supports these values, organizations can enhance job satisfaction and performance [@stillman2010]. Moreover, we found that job autonomy is the most important factor in mediating the relationship between free will beliefs and job satisfaction. When employees feel they have the freedom to make decisions and control their work processes, this autonomy reinforces their free will beliefs, leading to higher job satisfaction [@goezukara2016]. Our analysis shows that this applies directly to the workplace, since we found a robust effect of free will beliefs on job satisfaction, even when controlling for other psychological agency constructs, suggests a strong and direct relationship. As @clark2017 have found, belief in free will also alleviates distress and anxiety in regards to punishment and moral judgments, which can further positively impact satisfaction at work. This is crucial information, since it can help us effectively design targeted interventions with long-term benefits for employees and organizations [@feldman2017].

## Demographic Variable Effects
Our findings suggest that while there are slight differences in the median belief in free will among different age groups, these differences do not reach statistical significance. This outcome indicates that the belief in free will might be relatively stable across different stages of adult life, not significantly impacted by age-related factors such as life experience and generational values, but rather by cultural factors @feltz2009. Our findings also suggest that a slight differences based on sex, with women reporting higher beliefs in free will. This finding could reflect underlying gender-related socialization processes, where differing societal expectations and roles for men and women might influence their perceptions of autonomy and control [@baumeister1988]. However, even when found significant, this effect was marginal in both samples, suggesting that the influence of free will beliefs on job satisfaction might operate independently of these demographic factors. This independence has strong implications for potential interventions interventions aimed at enhancing employees' sense of autonomy and control could be broadly applicable and beneficial [@sheldon2001].

@@

## Future Directions
The next steps to take would be to further explore societal changes over time and individual differences in the value of free will, which would be possible due to the longitudinal nature of the WVS data. Further research couldalso  investigate the impact of emerging technologies and digital platforms on perceptions of free will, examining how online behavior and the emergence of artificial intelligence influences or reflects these beliefs. Additionally, interdisciplinary approaches combining psychology, sociology, and neuroscience could offer deeper insights into the biological underpinnings and social determinants of free will beliefs. Exploring these areas could significantly enhance our understanding of free will's role in modern society and individual psychology, and help us even more in designing appropriate interventions not only at the workspace but other spheres of our lives.

>>>>

Utilizing the same dataset and a similar methodology to @feldman2017, we aim to replicate their results in the context of the United States, a very heterogenous country that is characterized by a high importance of individualist, but also cultural diversity and socioeconomic inequality. While Feldman et al. did investigate the relationship between the belief in free will and job satisfaction in the United States, they did not integrate these results to the cross-cultural analysis, as the US were not included in the sample of 16 countries. This paper, therefore, seeks to fill this gap and reproduce these aspects of the Feldman et al. paper: (1) the relationship between the belief in free will and job satisfaction applies to the context of American adults working at Amazon Mechanical Turk (2) free will is a significant predictor of job satisfaction at the national level, using the World Values Survey. We will also consider whether there are any significant differences between the national-level data and the Amazon Mechanical Turk data. The estimand is the correlation between the belief in free will and job satisfaction. Utilizing Amazon Mechanical Turk and World Values Survey data, we found that belief in free will significantly correlates with job satisfaction both at individual and national levels. Additionally, we found job autonomy as a critical mediator in this relationship, while other agency constructs and demographic factors play a less important role in impacting workplace satisfaction across diverse American contexts. We also found American participants to be among the world's highest believers in free will, further emphasizing the importance of considering the local context.

We begin our paper with a closer look at the data source and measurement (@sec-data), followed by a selection of relevant results (@sec-results). Finally, we will conclude with a discussion of our findings to integrate our replication with the original paper by @feldman2017 and the broader literature on free will beliefs (@sec-discussion). We will also discuss potential ethical biases, limitations, and future directions. Our reproduction was conducted using the statistical programming language `R` [@citeR]. In the data analysis and visualization process, we also made use of the following `R` packages: `tidyverse` [@citetidyverse], `dplyr` [@citedplyr], `knitr` [@citeknitr], `kableExtra` [@citekExtra], `purrr` [@citepurrr], `psych` [@citepsych], `car` [@citecar], `MASS` [@citeMASS], and `countrycode` [@citecc].

@@

* Pride in Job: We expect this variable to positively correlate with job satisfaction and job autonomy, such that it can serve as some indication of internal validity and reliability. Not all waves of WVS data asked this question, but where possible we used the following question (C031):
  + How much pride, if any, do you take in the work that you do? (1 = "A great deal", 3 = "Little").
  + We also inverted this scale, such that 1 means "Little" and 3 means "A great deal", making it more consistent with the ordering logic of other variables used.

@@

kable(reliability_scores,
    booktabs = TRUE,
    linesep = "")
```
\newpage
## Data Characteristics {#sec-characteristics}

@@

kable(satisfaction_data_usa_summary,
    booktabs = TRUE,
    linesep = ""
)
```

@@

kable(wvs_data_usa_summary,
booktabs = TRUE,
    linesep = ""
)

@@

In exploring free will beliefs and their impact on job satisfaction in the context of American Amazon Mechanical Turk workers, our study replicates and extends the results of @feldman2017. As seen in @fig-corst2, most variables are significantly correlated with each other, with the highest correlations being unsurprisingly between the two measurements of Job satisfaction. Locus of control lacks any significant correlations, which together with its low reliability coefficient designates it as a less decisive control variable. The most important relationship, however, is between free will beliefs and job satisfaction: The belief in free will was significantly positively correlated with job satisfaction both at Time 1 (*r* = `r cor_fw_js1_correlation_coefficient`, *p* \<.001, CI \[`r cor_fw_js1_ci[1]`, `r cor_fw_js1_ci[2]`\]) and at Time 2 (r = `r cor_fw_js2_correlation_coefficient`, *p* \<.001, CI \[`r cor_fw_js2_ci[1]`, `r cor_fw_js1_ci[2]`\]). However, this correlation is not as strong as the one between free will beliefs and job autonomy (*r* = `r cor_fw_jaut1_correlation_coefficient`, *p* \<.001, CI \[`r cor_fw_jaut1_ci[1]`, `r cor_fw_jaut1_ci[2]`\]). This supports the claim that job autonomy may be the most important mediating variable between free will beliefs and job satisfaction. Moreover, a step-wise multiple regression on job satisfaction with all psychological agency constructs (full results in @sec-apx-stepwise) confirmed that the effect of free will beliefs on job satisfaction is robust and the strongest relative impact, even when controlling for other agency constructs (Time 1: *p* \< 0.034; Time 2: *p* \< 0.031).

@@

ggplot(satisfaction_data_usa_rel, aes(x = factor(Sex, labels = c("Male", "Female")), y = `Belief in Free Will (T1)`)) +
  geom_boxplot(fill = "lightblue") +
  labs(
       x = "Sex",
       y = "Belief in Free Will") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  + theme_minimal()
```

@@

## Effect of Free Will Beliefs
The main result of our analysis is that the belief in free will is sufficient to positively influence one's job satisfaction. Both the Amazon Mechanical Turk and WVS data sources showed significant positive correlations between these two psychological constructs. Free will beliefs are found to be  integral to psychological well-being and motivation, and are commonly linked to agency, autonomy, and responsibility [@wang2009]. By fostering an environment that supports these values, organizations can enhance job satisfaction and performance [@stillman2010]. Moreover, we found that job autonomy is the most important factor in mediating the relationship between free will beliefs and job satisfaction. When employees feel they have the freedom to make decisions and control their work processes, this autonomy reinforces their free will beliefs, leading to higher job satisfaction [@goezukara2016]. Our analysis shows that this applies directly to the workplace, since we found a robust effect of free will beliefs on job satisfaction, even when controlling for other psychological agency constructs, suggests a strong and direct relationship. As @clark2017 have found, belief in free will also alleviates distress and anxiety in regard to punishment and moral judgments, which can further positively impact satisfaction at work. This is crucial information, since it can help us effectively design targeted interventions with long-term benefits for employees and organizations [@feldman2017].

## Demographic Variable Effects
Our findings suggest that while there are slight differences in the median belief in free will among different age groups, these differences do not reach statistical significance. This outcome indicates that the belief in free will might be relatively stable across different stages of adult life, not significantly impacted by age-related factors such as life experience and generational values, but rather by cultural factors @feltz2009. Our findings also suggest that a slight difference based on sex, with women reporting higher beliefs in free will. This finding could reflect underlying gender-related socialization processes, where differing societal expectations and roles for men and women might influence their perceptions of autonomy and control [@baumeister1988]. However, even when found significant, this effect was marginal in both samples, suggesting that the influence of free will beliefs on job satisfaction might operate independently of these demographic factors. This independence has strong implications for potential interventions aimed at enhancing employees' sense of autonomy and control could be broadly applicable and beneficial [@sheldon2001].

@@

## Future Directions
The next steps to take would be to further explore societal changes over time and individual differences in the value of free will, which would be possible due to the longitudinal nature of the WVS data. Further research could also  investigate the impact of emerging technologies and digital platforms on perceptions of free will, examining how online behavior and the emergence of artificial intelligence influences or reflects these beliefs. Additionally, interdisciplinary approaches combining psychology, sociology, and neuroscience could offer deeper insights into the biological underpinnings and social determinants of free will beliefs. Exploring these areas could significantly enhance our understanding of free will's role in modern society and individual psychology, and help us even more in designing appropriate interventions not only at the workspace but other spheres of our lives.
